### 2025-07-14 paper/sections/experiments.qmd

1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

@@

Our key outcome of interest is improvement in explainability (Def. \ref{def-explainability}). Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time^[Additional metrics such as validity and redundancy are reported in the appendix.]. To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

We also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). These datasets are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a. Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2. Finally, for the convenience of illustration, we use the 10-class *MNIST* dataset [@lecun1998mnist].

@@

Our main quantitative results for the *MLP* are summarised in @tbl-main, which shows average outcomes along with bootstrapped standard errors for key metrics. The following example illustrates how to read @tbl-main.

@@

### Synthetic Example: Prediction of Credit Defaults

@fig-poc presents results for a linear classifier fitted to *LS*, which complies with the data assumptions in Proposition \ref{prp-mtblty}. The four panels show the outcomes for different training procedures: in panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT. For illustrative purposes, suppose the first feature represents *debt* (mutable) and the second feature represents *age* (immutable) of loan applicants seeking counterfactual explanations for moving to the target class: loan provided (orange).

@@

In panel (c) of @fig-poc, the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals. This result is quantified for the non-linear case in column three of @tbl-main, which shows the $\%$-reduction in costs averaged across valid counterfactuals. Once again, the impact of CT is statistically significant and substantial (14%). Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}. In the following, we present the results for all remaining datasets.

@@

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$. Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).

![Explanations for *MNIST* for BL (top) and CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.]. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

@@

Key performance metrics and bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Standard errors for accuracy are bootstrapped from the test set.
:::

>>>>

1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
2. To what extent does CT lead to more favorable AR outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection on CT?

@@

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

For predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

@@

Our main quantitative results for *MLP* models are summarised in @tbl-main, which presents average outcomes along with bootstrapped two standard errors. The following example motivates CT and illustrates how to read @tbl-main.

@@

### Synthetic Example: Prediction of Credit Card Defaults.

@fig-poc presents results for a linear classifier fitted to *LS* that complies with the data assumptions in Proposition \ref{prp-mtblty}. The four panels show the outcomes for different training procedures: in panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT. For illustrative purposes, suppose the first feature represents *debt* (mutable) and the second feature represents *age* (immutable) of loan applicants seeking counterfactual explanations for moving to the target class: loan provided (orange).

@@

In panel (c) of @fig-poc, the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These CEs are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals. This result is quantified for the non-linear case in column three of @tbl-main, which shows the $\%$-reduction in costs averaged across valid counterfactuals. Once again, the impact of CT is statistically significant and substantial (14%). Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}. Next, we present the results for all remaining datasets.

@@

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$. Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).

![Sample explanations for *MNIST* for BL (top) and CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

@@

Key performance metrics and bootstrapped *two* standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Standard errors for accuracy are bootstrapped from the test set.
:::

### 2025-07-14 paper/sections/intro.qmd

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery]. Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if learned representations correspond to meaningful, plausible explanations. Secondly, even if we resolve this challenge, it remains undecided how to ensure that machine learning models can *only* learn valuable explanations.

>>>>

Part of the problem is that the high degrees of freedom---high number of parameters estimated from data---provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]. As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery]. Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if learned representations correspond to meaningful, plausible explanations. Secondly, even if we resolve this challenge, it remains undecided how to ensure that machine learning models can *only* learn valuable explanations.
